---
title: "Basic use of Sinba model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic use of Sinba model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sinba)
```

The package sinba implements the **Sinba** (single birth of a trait) model.
The Sinba model is used to test the correlation
between two traits using a phylogenetic comparative method framework
originally developed by Pagel (1994).

# Correlated trait evolution

One of the main questions in evolutionary biology
is if two traits evolve independently
or in a correlated fashion.
For example,
suppose that we know the phylogenetic relationships of a clade,
and we observe four traits on the clade members,
*m*, *n*, *p*, and *q*:

```{r}
# Based on an example in section 7.2.2
# from the book of Revell & Harmon (2022)
# pp. 173-177.
library(phytools)
library(RColorBrewer)

# seed to make the example reproducible
set.seed(6)

tree <- pbtree(n = 26, tip.label = LETTERS)
m <- c(
  0, 0, 1, 1, 1, 1, 1, 0, 0, 0, 0, 1, 1, 1,
  1, 0, 0, 0, 1, 0, 1, 1, 0, 0, 0, 0
)
n <- m
p <- c(
  0, 0, 1, 0, 1, 0, 1, 0, 0, 0, 0, 0, 1, 0,
  1, 0, 0, 0, 1, 0, 1, 0, 0, 0, 0, 0
)
q <- rep(c(1, 0), 13)

plotTree.datamatrix(tree,
  data.frame(
    m = as.factor(setNames(m, LETTERS)),
    n = as.factor(setNames(n, LETTERS)),
    p = as.factor(setNames(p, LETTERS)),
    q = as.factor(setNames(q, LETTERS)),
    row.names = LETTERS
  ),
  fsize = 1, space = 0.2, xexp = 1.4,
  palettes = c("YlOrBr", "Blues", "Greens", "Purples")
)
```

It looks like *m* and *n* are correlated
(they have the same distribution),
*m* and *p* seem to be correlated
(*p* has only 1 if *m* is 1),
whereas *m* and *q* seem to be independent.

# Pagel's method

To represent the evolution of a trait in a phylogenetic tree,
we use a Markov model.
This model is formalized by a **Q matrix**
that contains the instantaneous transition rates between each state.
For example,
in a binary trait the Q matrix is:

$$
Q=\begin{bmatrix}
- & q_{0\to 1} \\
q_{1\to 0} & - \\
\end{bmatrix}
$$

Using the pruning algorithm of Felsenstein (1981),
it is possible to estimate the likelihood of a trait,
given a phylogenetic tree (with branch lengths)
and the model Q matrix.
If you got two *independent* traits,
then the full likelihood is the multiplication of the likelihood of both traits.

```{r}
# we do the fit of a single trait using PhyTools

fit_m <- fitMk(tree, as.factor(setNames(m, LETTERS)),
  model = "ARD", pi = "fitzjohn"
)
fit_n <- fitMk(tree, as.factor(setNames(n, LETTERS)),
  model = "ARD", pi = "fitzjohn"
)
fit_p <- fitMk(tree, as.factor(setNames(p, LETTERS)),
  model = "ARD", pi = "fitzjohn"
)
fit_q <- fitMk(tree, as.factor(setNames(q, LETTERS)),
  model = "ARD", pi = "fitzjohn"
)
fit_m$logLik
fit_n$logLik
fit_p$logLik
fit_q$logLik
```

But what happens if both traits are correlated?
In such a case,
the transformation between two states in a trait
is different given the state of the second trait.
Pagel (1994) solved that problem
by creating a Q matrix that combines the two traits.
The constraint of this new matrix
is that only a single transformation is allowed at a single time.
Then the Q matrix for both traits is:

$$
Q=\begin{bmatrix}
- & q_{00\to 01} & q_{00\to 10} & 0 & \\
q_{01\to 00} & - & 0 & q_{01\to 11} & \\
q_{10\to 00} & 0 & - & q_{10\to 11} & \\
0 & q_{11\to 01} & q_{11\to 10} & - & \\
\end{bmatrix}
$$

The advantage of this formalization
is that when trait changes are independent
(i.e., $q_{00\to 01} = q_{10\to 11}$,
$q_{00\to 10} = q_{01\to 11}$,
$q_{01\to 00} = q_{11\to 10}$,
and $q_{10\to 00} = q_{11\to 01}$),
the resulting likelihood is identical to modeling both traits independently.
Then we can test for correlation between the two traits
using the likelihood of the Q matrix
and the number of free parameters,
using,
for example, the Akaike information criterion
(AIC, Akaike, 1974).

In the `sinba` package,
the Pagel model is run with the function `fit_pagel()`,
and the models,
either independent evolution or correlated evolution,
are defined with the "IND" (the default) and "CORR" models:

```{r}
# in sinba the data is a data frame
# in which the first column contains the name of the taxa
# and other remaining columns the data.
data_mn <- data.frame(taxa = LETTERS, m = m, n = n)
data_mp <- data.frame(taxa = LETTERS, m = m, p = p)
data_mq <- data.frame(taxa = LETTERS, m = m, q = q)

# --- m vs n
pagel_mn_ind <- fit_pagel(tree, data_mn)

# compare the likelihood of both traits calculated independently
# (calculated with PhyTools)
# with the likelihood from the independent Pagel's model
# (calculated in sinba package)
fit_m$logLik + fit_n$logLik
pagel_mn_ind$logLik

# compare the independent and correlated model
# using AIC
pagel_mn_corr <- fit_pagel(tree, data_mn, model = new_model("CORR"))
aic_mn <- setNames(
  c(
    AIC(pagel_mn_ind),
    AIC(pagel_mn_corr)
  ),
  c("Pagel-IND", "Pagel-CORR")
)
aic_mn

# --- m vs p
pagel_mp_ind <- fit_pagel(tree, data_mp)

# compare the likelihood of both traits calculated independently
# (calculated with PhyTools)
# with the likelihood from the independent Pagel's model
# (calculated in sinba package)
fit_m$logLik + fit_p$logLik
pagel_mp_ind$logLik

# compare the independent and correlated model
# using AIC
pagel_mp_corr <- fit_pagel(tree, data_mp, model = new_model("CORR"))
aic_mp <- setNames(
  c(
    AIC(pagel_mp_ind),
    AIC(pagel_mp_corr)
  ),
  c("Pagel-IND", "Pagel-CORR")
)
aic_mp

# --- m vs q
pagel_mq_ind <- fit_pagel(tree, data_mq)

# compare the likelihood of both traits calculated independently
# (calculated with PhyTools)
# with the likelihood from the independent Pagel's model
# (calculated in sinba package)
fit_m$logLik + fit_q$logLik
pagel_mq_ind$logLik

# compare the independent and correlated model
# using AIC
pagel_mq_corr <- fit_pagel(tree, data_mq, model = new_model("CORR"))
aic_mq <- setNames(
  c(
    AIC(pagel_mq_ind),
    AIC(pagel_mq_corr)
  ),
  c("Pagel-IND", "Pagel-CORR")
)
aic_mq
```

As we suspected,
*m* and *n*,
and *m* and *p*,
evolve in a correlated fashion,
whereas *m* and *q* evolve independently.

# Darwin's dilemma

While the logic of Pagel's method is straightforward,
Maddison & FitzJohn (2015)
found that in some cases,
in particular,
when one of the traits has a single origin,
the results of Pagel's method are misleading.

```{r}
# An example from the book of Revell & Harmon (2022)
# section 7.2.2,
# pp. 173-177.

x <- c(0, 0, rep(1, 12), rep(0, 12))
y <- x
z <- c(0, 0, rep(c(0, 1), 6), rep(0, 12))

plotTree.datamatrix(tree,
  data.frame(
    x = as.factor(setNames(x, LETTERS)),
    y = as.factor(setNames(y, LETTERS)),
    z = as.factor(setNames(z, LETTERS)),
    row.names = LETTERS
  ),
  fsize = 1, space = 0.2, xexp = 1.4,
  palettes = c("YlOrBr", "Greens", "Purples")
)
```

In this example,
*x* and *y* have identical distributions,
but in contrast with *m* and *n*,
they only have a single origin.
This scenario is known as "*Darwin's dilemma*."
In the case of *z*,
it seems that its evolutionary rate sped up in the same clade
in which *x* originated.
This scenario is known as the "*unreplicated burst*."

The expectation is that in both scenarios,
the preferred model should be the independent model,
as we do not have enough evidence
to suggest a trait correlation
beyond that they happen to originate in the same clade.

```{r}
# Darwin's dilemma
data_xy <- data.frame(LETTERS, x, y)

pagel_xy_ind <- fit_pagel(tree, data_xy, model = new_model("IND"))
pagel_xy_corr <- fit_pagel(tree, data_xy, model = new_model("CORR"))
aic_xy <- setNames(
  c(
    AIC(pagel_xy_ind),
    AIC(pagel_xy_corr)
  ),
  c("Pagel-IND", "Pagel-CORR")
)
aic_xy

# Unreplicated burst
data_xz <- data.frame(LETTERS, x, z)

pagel_xz_ind <- fit_pagel(tree, data_xz, model = new_model("IND"))
pagel_xz_corr <- fit_pagel(tree, data_xz, model = new_model("CORR"))

aic_xz <- setNames(
  c(
    AIC(pagel_xz_ind),
    AIC(pagel_xz_corr)
  ),
  c("Pagel-IND", "Pagel-CORR")
)
aic_xz
```

But contrary to our expectations,
in both scenarios,
the preferred model is the correlated model.

# The Sinba model

The main problem with Darwin's dilemma
and the unreplicated burst scenario
is the presence of many "00" tips,
which suggests that whenever *x* is "0",
*y* should also be "0".
But this could be explained better
if it just happened that the birth of state "1"
in both traits
just happened at the same branch,
so the Q matrix cannot model the evolution in basal
or independent branches
because there is no process at all.
In the Sinba model
(for the *single birth of a trait*),
a trait starts as *inactive*:
no transitions happen.
Then there is the birth of one of the traits,
so the process is *semi-active*
(i.e., only one trait can change).
Finally,
there is the birth of the second trait,
so the process is a full process,
or *active*.

Apart from the parameters of the Q matrix,
the Sinba model has two additional parameters:
the location of each birth event.
It also requires the starting state after each birth event
(by default,
the value is set as 1.0 for the trait labeled as 1).

In the `sinba package` the function to fit a pair of traits
to the sinba model is called `fit_sinba()`.
By default,
it uses the independent model,
but as for the Pagel's model,
we can define a correlated model as well:

```{r}
# --- x and y (Darwin's scenario)

# Sinba model for the Darwin's scenario with independent evolution
sinba_xy_ind <- fit_sinba(tree, data_xy)
# Sinba model for the Darwin's scenario with correlated evolution
sinba_xy_corr <- fit_sinba(tree, data_xy, model = new_model("CORR"))

# compare AIC values from Pagel's models
# and Sinba models
aic_xy <- setNames(
  c(
    AIC(pagel_xy_ind),
    AIC(pagel_xy_corr),
    AIC(sinba_xy_ind),
    AIC(sinba_xy_corr)
  ),
  c("Pagel-IND", "Pagel-CORR", "Sinba-IND", "Sinba-CORR")
)
aic_xy

# --- x and z (unreplicated burst scenario)

# Sinba model for the  unreplicated burst scenario with independent evolution
sinba_xz_ind <- fit_sinba(tree, data_xz, model = new_model("IND"))
# Sinba model for the  unreplicated burst scenario with independent evolution
sinba_xz_corr <- fit_sinba(tree, data_xz, model = new_model("CORR"))

# compare AIC values from Pagel's models
# and Sinba models
aic_xz <- setNames(
  c(
    AIC(pagel_xz_ind),
    AIC(pagel_xz_corr),
    AIC(sinba_xz_ind),
    AIC(sinba_xz_corr)
  ),
  c("Pagel-IND", "Pagel-CORR", "Sinba-IND", "Sinba-CORR")
)
aic_xz
```

For both scenarios,
the sinba model with independent evolution
has the best fit
among the examined alternatives
(Pagel and Sinba models
with independent and correlated evolution).

# Cited literature

Akaike, H.
(1974).
A new look at the statistical model identification.
*IEEE transactions on automatic control*,
**19**, 716-723.
DOI: [10.1109/TAC.1974.1100705](https://doi.org/10.1109/TAC.1974.1100705).

Felsenstein, J.
(1981).
Evolutionary trees from DNA sequences: a maximum likelihood approach.
*Journal of molecular evolution*,
**17**, 368-376.
DOI: [10.1007/BF01734359](https://doi.org/10.1007/BF01734359).

Maddison, W. P., & FitzJohn, R. G.
(2015).
The unsolved challenge to phylogenetic correlation tests for categorical characters.
*Systematic Biology*,
**64**, 127-136.
DOI: [10.1093/sysbio/syu070](https://doi.org/10.1093/sysbio/syu070).

Pagel, M.
(1994).
Detecting correlated evolution on phylogenies: a general method for the comparative analysis of discrete characters.
*Proceedings of the Royal Society of London. Series B: Biological Sciences*,
**255**, 37-45.
DOI: [10.1098/rspb.1994.0006](https://doi.org/10.1098/rspb.1994.0006).

Revell, L. J., & Harmon, L. J.
(2022).
*Phylogenetic Comparative Methods in R*,
Princeton Univ.,
Princeton (U.S.A.).