---
title: "A tour of sinba package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A tour of sinba package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sinba)
```

In this vignette
we will explore the functionality of the `sinba` package.
It is expected that you have some familiarity with
`vignette("Basic use of Sinba model")`
and `vignette("How to make custom models")`.

# Input data

The main data sources in sinba package are a *phylogenetic tree*,
a `ape::phylo` object as use in packages such as `package("ape")`
(Paradis & Schliep, 2019)
and `package("phytools")`
(Revell, 2024),
and the *data of the observed tips* in a data frame.
The first column of the data frame
contains the name of the tips of the tree,
and the following columns contain the observed state of the data
(either 0 or 1).

```{r}
library(phytools)

# seed to make the example reproducible
set.seed(6)

# a phylogenetic tree
tree <- pbtree(n = 26, tip.label = LETTERS)

x <- c(
  1, 1, 1, 1, 0, 1, 1, 1, 1, 0, 1, 1, 1, 1,
  rep(0, 12)
)
y <- c(
  0, 0, 1, 1, 1, 1, 0, 0, 1, 1, 1, 0, 1, 1,
  rep(0, 12)
)

# observed data
data_xy <- data.frame(taxa = LETTERS, x = x, y = y)
```

Because the tree is a `ape::phylo` object,
we can use `phytools` to plot the data:

```{r}
library(RColorBrewer)

plotTree.datamatrix(tree,
  data.frame(
    x = as.factor(setNames(data_xy$x, data_xy$taxa)),
    y = as.factor(setNames(data_xy$y, data_xy$taxa)),
    row.names = data_xy$taxa
  ),
  fsize = 1, space = 0.2, xexp = 1.4,
  palettes = c("Blues", "Greens")
)
```

# Fitting models

The main usage of the `sinba` package
is to fit models
to detect the correlation between two traits
using the *sinba* ("single birth of a trait") model.

## A single trait

Before attempting to test for correlation,
here we will compare the traditional *Mk* model
(Pagel, 1994;
Lewis, 2001)
with a *sinba* model for a single trait.
The main function is `fit_sinba()`,
as this function by default assumes a model with two traits,
we should explicitly define a model
with a single trait with the function `new_model()`
using the parameter `traits`.
As a new model with a single trait will be by default an "*ARD*" model,
we need to explicitly define it to be an "*ER*" model.
We need to define a *reset vector*.
This vector indicates what happened at the birth event.
By default,
the functions in the `sinba` package
assume that the state of the trait after a birth
is always in the state 1.
If you do not know the initial state,
you can set it to (0.5, 0.5).
The function parameter to set the reset vector is `pi_x`.
By default,
fit functions in the `sinba` package
use the root prior defined by FitzJohn et al. (2009),
that is,
scaling the likelihood at the root
with the normalized likelihood values.
Specific root priors can be defined with the `root` parameter.

```{r}
# fit a single trait using PhyTools
# an equal rates model
fit_x_er <- fitMk(tree, as.factor(setNames(data_xy$x, data_xy$taxa)),
  model = "ER", pi = "fitzjohn"
)
# an all rates different model
fit_x_ard <- fitMk(tree, as.factor(setNames(data_xy$x, data_xy$taxa)),
  model = "ARD", pi = "fitzjohn"
)

# now we fit the sinba model with a single trait
# an equal rates different model
sinba_x_er <- fit_sinba(
  tree, data_xy,
  # we set the model to ER model
  # with a single trait.
  model = new_model(model = "ER", traits = 1),
  # As this is the default
  # we can safely ignore the pi_x parameter.
  # We put it here just for completeness
  pi_x = c(0, 1),
  # As we use the default
  # (the FitzJohn prior)
  # we can safely ignore the root parameter.
  # We put it here just for completeness
  root = c(0, 0)
)
# an all rates different model
sinba_x_ard <- fit_sinba(tree, data_xy,
  # the default model for a single trait is the ARD model
  model = new_model(traits = 1),
  # As this is the default
  # we can safely ignore the pi_x parameter.
  # We put it here just for completeness
  pi_x = c(0, 1),
  # As we use the default
  # (the FitzJohn prior)
  # we can safely ignore the root parameter.
  # We put it here just for completeness
  root = c(0, 0)
)

# the fit results can be printed with useful information
# about the results
sinba_x_ard

# as fit objects in phytools and sinba implement the AIC generic
# we can easily compare the models:
aic_single <- setNames(
  c(
    AIC(fit_x_er),
    AIC(fit_x_ard),
    AIC(sinba_x_er),
    AIC(sinba_x_ard)
  ),
  c("Mk-ER", "Mk-ARD", "sinba-ER", "sinba-ARD")
)
aic_single
```

As an exercise,
do the same analysis,
but for the second trait.

## Testing correlation between two traits

The main use of the `sinba` package
is to test the correlation between two traits
(Pagel, 1994).
To fit the data to a sinba model
we use the `fit_sinba()` function.
If we want to fit the data to a traditional Pagel model,
we use the function `fit_pagel()`.
In both functions we need to define a model
(the default is the independent model).
As we are using default "canned" models,
we can use the function `new_model()`
with the parameter `model` set as "*CORR*".
In the case of the sinba model,
we need to define the reset vector for both traits.
By default,
in `sinba` package this vector is set to be 1.0
for the state 1 in both *x* and *y* traits.

```{r}
# Pagel model
pagel_xy_ind <- fit_pagel(tree, data_xy)
pagel_xy_corr <- fit_pagel(tree, data_xy, model = new_model("CORR"))

# print the results
pagel_xy_ind
pagel_xy_corr

# sinba model
sinba_xy_ind <- fit_sinba(tree, data_xy,
  # As this is the default
  # we can safely ignore both the the pi_x and pi_y parameters.
  # We put them here just for completeness
  pi_x = c(0, 1), pi_y = c(0, 1),
  # As we use the default
  # (the FitzJohn prior)
  # we can safely ignore the root parameter.
  # We put it here just for completeness
  root = c(0, 0, 0, 0)
)
sinba_xy_corr <- fit_sinba(tree, data_xy,
  model = new_model("CORR"),
  # as this is the default
  # we can safely ignore both the the pi_x and pi_y parameters.
  # We put them here just for completeness
  pi_x = c(0, 1), pi_y = c(0, 1),
  # as we use the default
  # (the FitzJohn prior)
  # we can safely ignore the root parameter.
  # We put it here just for completeness
  root = c(0, 0, 0, 0)
)

# print the results
sinba_xy_ind
sinba_xy_corr

# compare AIC values from Pagel's models
# and Sinba models
aic_xy <- setNames(
  c(
    AIC(pagel_xy_ind),
    AIC(pagel_xy_corr),
    AIC(sinba_xy_ind),
    AIC(sinba_xy_corr)
  ),
  c("Pagel-IND", "Pagel-CORR", "Sinba-IND", "Sinba-CORR")
)
aic_xy
```

## Other trait models

The `sinba` package includes several canned models,
and also it is possible to edit model parameters,
so the model universe is quite flexible.
See the `vignette("How to make custom models")`
for more information.
These models can be used by both `fit_sinba()`
and `fit_pagel()` functions.

Here,
we want to also fit the equal rates model ("*ER*")
and the symmetric model ("*SYM*").
We also use a custom model
that has equal rates for trait *x*
and all rates different for trait *y*.

```{r}
# Equal Rates model with sinba
fit_sinba(tree, data_xy, new_model("ER"),
  # as this is the default
  # we can safely ignore both the the pi_x and pi_y parameters.
  # We put them here just for completeness.
  pi_x = c(0, 1), pi_y = c(0, 1),
  # as we use the default
  # (the FitzJohn prior)
  # we can safely ignore the root parameter.
  # We put it here just for completeness
  root = c(0, 0, 0, 0)
)

# Symmetric model with sinba
fit_sinba(tree, data_xy, new_model("SYM"),
  # as this is the default
  # we can safely ignore both the the pi_x and pi_y parameters.
  # We put them here just for completeness
  pi_x = c(0, 1), pi_y = c(0, 1),
  # as we use the default
  # (the FitzJohn prior)
  # we can safely ignore the root parameter.
  # We put it here just for completeness
  root = c(0, 0, 0, 0)
)

# a custom model
m <- new_model()
m <- model_equate(m, c(2, 4))
m

# fit the custom model with sinba
fit_sinba(tree, data_xy, m,
  # as this is the default
  # we can safely ignore both the the pi_x and pi_y parameters.
  # We put them here just for completeness
  pi_x = c(0, 1), pi_y = c(0, 1),
  # as we use the default
  # (the FitzJohn prior)
  # we can safely ignore the root parameter.
  # We put it here just for completeness
  root = c(0, 0, 0, 0)
)
```

## Alternative birth models

Suppose we got the trait *w*
and we want to couple it with the previous trait *x*:

```{r}
w <- c(
  1, 0, 1, 1, 1, 1, 0, 1, 0, 0, 1, 1, 0, 1,
  rep(0, 12)
)

data_xw <- data.frame(taxa = LETTERS, x = x, y = w)

# plot the data
plotTree.datamatrix(tree,
  data.frame(
    x = as.factor(setNames(data_xw$x, data_xw$taxa)),
    w = as.factor(setNames(data_xw$y, data_xw$taxa)),
    row.names = data_xw$taxa
  ),
  fsize = 1, space = 0.2, xexp = 1.4,
  palettes = c("Blues", "Greens")
)

# fit sinba model
fit_sinba(tree, data_xw,
  # as this is the default
  # we can safely ignore both the the pi_x and pi_y parameters.
  # We put them here just for completeness
  pi_x = c(0, 1), pi_y = c(0, 1),
  # as we use the default
  # (the FitzJohn prior)
  # we can safely ignore the root parameter.
  # We put it here just for completeness
  root = c(0, 0, 0, 0)
)
```

Note that both traits are born at the same node,
so maybe we can fit a model in which both traits are born simultaneously,
and then we can drop one of the birth event parameters.
This can be done with the function `fit_simultaneous()`.
In this case,
our reset vector has four elements
(as any combination of the four states are possible;
they are 00, 01, 10, and 11).
The default is that the trait starts at 11,
but can be changed with the `pi_xy` parameter.

```{r}
# fit sinba model with a simultaneous birth
fit_simultaneous(tree, data_xw,
  # as this is the default
  # we can safely ignore the pi_xy parameter.
  # We put it here just for completeness
  pi_xy = c(0, 0, 0, 1),
  # as we use the default
  # (the FitzJohn prior)
  # we can safely ignore the root parameter.
  # We put it here just for completeness
  root = c(0, 0, 0, 0)
)
```

In the following example,
we got the trait *z*,
and we want to model it simultaneously with *x*:

```{r}
z <- c(
  0, 1, 1, 1, 1, 0, 1, 1, 1, 0, 1, 0, 1, 1,
  1, 0, 1, 1, 0, 1, 1, 1, 0, 0, 1, 1
)

data_xz <- data.frame(taxa = LETTERS, x = x, y = z)

# plot the data
plotTree.datamatrix(tree,
  data.frame(
    x = as.factor(setNames(data_xz$x, data_xz$taxa)),
    z = as.factor(setNames(data_xz$y, data_xz$taxa)),
    row.names = data_xz$taxa
  ),
  fsize = 1, space = 0.2, xexp = 1.4,
  palettes = c("Blues", "Greens")
)

# fit sinba model
fit_sinba(tree, data_xz,
  # as this is the default
  # we can safely ignore both the the pi_x and pi_y parameters.
  # We put them here just for completeness
  pi_x = c(0, 1), pi_y = c(0, 1),
  # as we use the default
  # (the FitzJohn prior)
  # we can safely ignore the root parameter.
  # We put it here just for completeness
  root = c(0, 0, 0, 0)
)
```

Note that the second trait is born at the root
(or at any point before the root node),
so instead of fitting two birth events,
we can fit just a single birth event
and leave the second event as a birth fixed at the root.
The function to do that is `fit_mixed()`.
In this case,
we need a single reset vector,
`pi_trait`,
which by default assumes that the trait is born at state 1.
We also need to define the trait that is a free parameter
(the parameter is `trait` and by default,
as in this case,
is *x*,
the first trait):

```{r}
# fit sinba with a free birth and a birth fixed at the root
fit_mixed(tree, data_xz,
  # this is the default
  # so we can safely ignore trait parameter.
  # We put it here for completeness
  trait = "x",
  # as this is the default
  # we can safely ignore the pi_trait parameter.
  # We put it here just for completeness
  pi_trait = c(0, 1),
  # as we use the default
  # (the FitzJohn prior)
  # we can safely ignore the root parameter.
  # We put it here just for completeness
  root = c(0, 0, 0, 0)
)
```

## Hidden models

Package `sinba` supports the definition of hidden models
(Marazzi et al., 2012;
Beaulieu et al., 2013).
In this example,
we observe the trait *h*
and want to model it in conjunction with trait *x*:

```{r}
h <- c(
  1, 1, 0, 1, 1, 1, 1, 1, 0, 1, 0, 1, 0, 1,
  rep(0, 12)
)

data_xh <- data.frame(taxa = LETTERS, x = x, y = h)

# plot the data
plotTree.datamatrix(tree,
  data.frame(
    x = as.factor(setNames(data_xh$x, data_xh$taxa)),
    h = as.factor(setNames(data_xh$y, data_xh$taxa)),
    row.names = data_xh$taxa
  ),
  fsize = 1, space = 0.2, xexp = 1.4,
  palettes = c("Blues", "Greens")
)

# fit sinba model
fit_sinba(tree, data_xh,
  # as this is the default
  # we can safely ignore both the the pi_x and pi_y parameters.
  # We put them here just for completeness
  pi_x = c(0, 1), pi_y = c(0, 1),
  # as we use the default
  # (the FitzJohn prior)
  # we can safely ignore the root parameter.
  # We put it here just for completeness
  root = c(0, 0, 0, 0)
)
```

Now we create a hidden model with two rates
and set that any rate change
should have the same rate:

```{r}
# a hidden model with two rates
hm <- new_rates_model(new_model("IND"), rates = c("a", "b"))
hm <- model_equate_trans_rates(hm)
hm
```

Now we fit the hidden model.
While the state space in hidden models is larger,
this space is unobserved,
so the reset vector and the root prior are defined
in the same way as in a non-hidden model
(just the values for 0 and 1,
in the case of the reset vector,
and the values of 00, 01, 10, and 11 for the root prior).
The functions in sinba
will automatically expand the vectors
taking into account the hidden states,
dividing the probability of each state
equally among hidden states
(for example, the vector
(0.20, 0.80),
in which both trait states have two hidden states,
will be expanded to have a prior of 0.1
for both hidden states corresponding to 0
and 0.4 for both hidden states corresponding to 1).

```{r}
# fit sinba model using the hidden model
fit_sinba(tree, data_xh,
  model = hm,
  # as this is the default
  # we can safely ignore both the the pi_x and pi_y parameters.
  # We put them here just for completeness
  pi_x = c(0, 1), pi_y = c(0, 1),
  # as we use the default
  # (the FitzJohn prior)
  # we can safely ignore the root parameter.
  # We put it here just for completeness
  root = c(0, 0, 0, 0)
)
```

In this example,
the hidden model does not improve
the simpler single-rate model.
That is expected in our tiny dataset.
In more realistic (bigger) data sets,
sometimes hidden models offer a considerably better explanation.

## Fixed values

Most of the time,
we want to infer the parameters of the model.
But in some cases,
we already have some fixed values for those parameters
(maybe from previous research
or just to see what if…?).

If we know the birth locations
and want to know which is the best Q matrix,
we can use the function `fit_fixed_births()`.

```{r}
# birth list
bb <- list(
  # trait x
  list(
    node = 28, # we use the node labels to identify nodes
    age = 0 # birth at the beginning of the branch
  ),
  # trait y
  list(
    node = 29,
    age = 0
  )
)

# fit sinba with fixed births
fit_fixed_births(tree, data_xy, bb,
  # as this is the default
  # we can safely ignore both the the pi_x and pi_y parameters.
  # We put them here just for completeness
  pi_x = c(0, 1), pi_y = c(0, 1),
  # as we use the default
  # (the FitzJohn prior)
  # we can safely ignore the root parameter.
  # We put it here just for completeness
  root = c(0, 0, 0, 0)
)
```

Maybe we know the Q matrix,
and we want to fit the birth locations.
In that case we use `fit_fixed_matrix()`.
Note that if you are using a hidden model,
you also need to include the model definition
(the `model` parameter).

```{r}
# Q matrix
q <- matrix(c(
  0,   5e-10, 5e-10, 0,
  0.6, 0,     0,     5e-10,
  0.4, 0,     0,     5e-10,
  0,   0.4,   0.6,   0
), byrow = TRUE, nrow = 4)

# fit sinba with fixed Q
fit_fixed_matrix(tree, data_xy, q,
  # the model itself has no influence,
  # it is just used to connect the observed states
  # with the matrix IDs,
  # which is need for hidden models.
  # By default,
  # it uses the independent model.
  model = new_model("IND"),
  # as this is the default
  # we can safely ignore both the the pi_x and pi_y parameters.
  # We put them here just for completeness
  pi_x = c(0, 1), pi_y = c(0, 1),
  # as we use the default
  # (the FitzJohn prior)
  # we can safely ignore the root parameter
  # We put it here just for completeness.
  root = c(0, 0, 0, 0)
)
```

If you have both the birth events
and the Q matrix,
you can estimate the likelihood using the `fixed_sinba()` function.

```{r}
# fit sinba with fixed births and Q matrix
fixed_sinba(tree, data_xy, q, bb,
  # the model itself has no influence,
  # it is just used to connect the observed states
  # with the matrix IDs,
  # which is need for hidden models.
  # By default,
  # it uses the independent model.
  model = new_model("IND"),
  # as this is the default
  # we can safely ignore both the the pi_x and pi_y parameters.
  # We put them here just for completeness
  pi_x = c(0, 1), pi_y = c(0, 1),
  # as we use the default
  # (the FitzJohn prior)
  # we can safely ignore the root parameter.
  # We put it here just for completeness
  root = c(0, 0, 0, 0)
)
```

# Stochastic mapping

With `sinba` package you can make stochastic maps
(Nielsen, 2002)
using the sinba model.
The function to do that is `map_sinba()`.
The resulting reconstruction
is an object of type `phytools::multiSimmap`,
so you can use `package("phytools")`
to manipulate it
(for example,
for plotting).

```{r}
# to make stochastic mappings
# you need a fitted sinba object
# (e.g. the results of `fit_sinba()`)
# and the number of simulations
# (default is 100).
mp <- map_sinba(sinba_xy_ind, 100)

# plot the first map
plot(mp[[1]])
```

Note that our single follows the expected pattern:
the traits do not change at all in the *inactive* part,
and only the first trait changes in the *semi-active* part.

```{r}
# make a summary (using phytools)
sm <- summary(mp)
sm

# plot the summary
plot(sm)
```

As can be seen from our 100 stochastic maps,
all the simulated histories are constrained
by our birth event inferences.

# Simulations

The `sinba` package also includes some facilities
to simulate data under the sinba model.
The function for simulations is `sim_sinba()`.

## Maddison & FitzJohn scenarios

The most simple simulations are based on Maddison & FitzJohn (2015) scenarios.
These are canned models that are built in an algorithmic way;
that is,
there is not a specific underlying model to generate the data.
But they are useful,
because they reflect situations
for which we have an intuitive idea
on how the methods should react.

```{r}
# a simulated tree with 50 tips
sim_tree <- pbtree(n = 50)

# simulate scenarios
darwin <- sim_sinba(sim_tree, scenario = "darwin")
unrep <- sim_sinba(sim_tree, scenario = "unrep")
codis <- sim_sinba(sim_tree, scenario = "codis")
burst <- sim_sinba(sim_tree, scenario = "burst")

# plot the darwin scenario
plotTree.datamatrix(sim_tree,
  data.frame(
    x = as.factor(setNames(darwin$data$x, darwin$data$t.tip)),
    y = as.factor(setNames(darwin$data$y, darwin$data$t.tip)),
    row.names = darwin$data$t.tip
  ),
  ftype = "off", space = 0.2, xexp = 1.4,
  palettes = c("Blues", "Greens")
)

# plot the unreplicated burst scenario
plotTree.datamatrix(sim_tree,
  data.frame(
    x = as.factor(setNames(unrep$data$x, unrep$data$t.tip)),
    y = as.factor(setNames(unrep$data$y, unrep$data$t.tip)),
    row.names = unrep$data$t.tip
  ),
  ftype = "off", space = 0.2, xexp = 1.4,
  palettes = c("Blues", "Greens")
)

# plot the codistribution scenario
plotTree.datamatrix(sim_tree,
  data.frame(
    x = as.factor(setNames(codis$data$x, codis$data$t.tip)),
    y = as.factor(setNames(codis$data$y, codis$data$t.tip)),
    row.names = codis$data$t.tip
  ),
  ftype = "off", space = 0.2, xexp = 1.4,
  palettes = c("Blues", "Greens")
)

# plot the replicated burst scenario
plotTree.datamatrix(sim_tree,
  data.frame(
    x = as.factor(setNames(burst$data$x, burst$data$t.tip)),
    y = as.factor(setNames(burst$data$y, burst$data$t.tip)),
    row.names = burst$data$t.tip
  ),
  ftype = "off", space = 0.2, xexp = 1.4,
  palettes = c("Blues", "Greens")
)
```

As an exercise,
fit the Pagel and sinba models to the simulated scenarios.

## Full simulation

A complete simulation under the sinba model
requires that we define all the model parameters,
that is,
the Q matrix and the birth events.
Simulated data sets always start at the 00 state.

```{r}
# Q matrix
qq <- matrix(c(
  0, 0.5, 0.2, 0,
  0.3, 0, 0, 0.2,
  0.4, 0, 0, 0.5,
  0, 0.4, 0.3, 0
), byrow = TRUE, nrow = 4)

# birth list
bb <- list(
  # trait x
  list(
    node = 28, # we use the node labels to identify nodes
    age = 0 # birth at the beginning of the branch
  ),
  # trait y
  list(
    node = 31,
    age = 0
  )
)

# make the simulation
sim_data <- sim_sinba(tree,
  rate_mat = qq, births = bb,
  # the model itself has no influence,
  # it is just used to connect the observed states
  # with the matrix IDs,
  # which is need for hidden models.
  # By default,
  # it uses the independent model.
  model = new_model("IND"),
  # as this is the default
  # we can safely ignore both the the pi_x and pi_y parameters.
  # We put them here just for completeness
  pi_x = c(0, 1), pi_y = c(0, 1),
)

# plot the simulated data
plotTree.datamatrix(tree,
  data.frame(
    x = as.factor(setNames(sim_data$data$x, sim_data$data$t.tip)),
    y = as.factor(setNames(sim_data$data$y, sim_data$data$t.tip)),
    row.names = LETTERS
  ),
  fsize = 1, space = 0.2, xexp = 1.4,
  palettes = c("Blues", "Greens")
)

# if birds are undefined
# it select two valid births at random
sim_data_50 <- sim_sinba(sim_tree,
  rate_mat = qq,
  # the model itself has no influence,
  # it is just used to connect the observed states
  # with the matrix IDs,
  # which is need for hidden models.
  # By default,
  # it uses the independent model.
  model = new_model("IND"),
  # as this is the default
  # we can safely ignore both the the pi_x and pi_y parameters.
  # We put them here just for completeness
  pi_x = c(0, 1), pi_y = c(0, 1)
)

# plot the simulated data
plotTree.datamatrix(sim_tree,
  data.frame(
    x = as.factor(setNames(sim_data_50$data$x, sim_data_50$data$t.tip)),
    y = as.factor(setNames(sim_data_50$data$y, sim_data_50$data$t.tip)),
    row.names = sim_data_50$data$t.tip
  ),
  ftype = "off", space = 0.2, xexp = 1.4,
  palettes = c("Blues", "Greens")
)
```

# Cited literature

Beaulieu, J. M., O'Meara, B. C., & Donoghue, M. J.
(2013).
Identifying hidden rate changes in the evolution of a binary morphological character: the evolution of plant habit in campanulid angiosperms.
*Systematic Biology*,
**62**, 725-737.
DOI: [10.1093/sysbio/syt034](https://doi.org/10.1093/sysbio/syt034).

FitzJohn, R. G., Maddison, W. P., & Otto, S. P.
(2009).
Estimating trait-dependent speciation and extinction rates from incompletely resolved phylogenies.
*Systematic Biology*,
**58**, 595-611.
DOI: [10.1093/sysbio/syp067](https://doi.org/10.1093/sysbio/syp067).

Lewis, P. O.
(2001).
A likelihood approach to estimating phylogeny from discrete morphological character data.
*Systematic biology*,
**50**, 913-925.
DOI: [10.1080/106351501753462876](https://doi.org/10.1080/106351501753462876).

Maddison, W. P., & FitzJohn, R. G.
(2015).
The unsolved challenge to phylogenetic correlation tests for categorical characters.
*Systematic Biology*,
**64**, 127-136.
DOI: [10.1093/sysbio/syu070](https://doi.org/10.1093/sysbio/syu070).

Marazzi, B., Ané, C., Simon, M. F., Delgado-Salinas, A., Luckow, M., & Sanderson, M. J.
(2012).
Locating evolutionary precursors on a phylogenetic tree.
*Evolution*,
**66**, 3918-3930.
DOI: [10.1111/j.1558-5646.2012.01720.x](https://doi.org/10.1111/j.1558-5646.2012.01720.x).

Nielsen, R.
(2002).
Mapping mutations on phylogenies.
*Systematic Biology*,
**51**, 729-739.
DOI: [10.1080/10635150290102393](https://doi.org/10.1080/10635150290102393).

Pagel, M.
(1994).
Detecting correlated evolution on phylogenies: a general method for the comparative analysis of discrete characters.
*Proceedings of the Royal Society of London. Series B: Biological Sciences*,
**255**, 37-45.
DOI: [10.1098/rspb.1994.0006](https://doi.org/10.1098/rspb.1994.0006).

Paradis, E., Schliep, K.
(2019).
ape 5.0: an environment for modern phylogenetics and evolutionary analyses in R.
*Bioinformatics*,
**35**, 526-528.
DOI: [10.1093/bioinformatics/bty633](https:://doi.org/10.1093/bioinformatics/bty633).

Revell, L.
(2024).
phytools 2.0: an updated R ecosystem for phylogenetic comparative methods (and other things).
*PeerJ*,
**12**, e16505.
DOI: [10.7717/peerj.16505](https://doi.org/10.7717/peerj.16505).